---
title: "Examples of using CPCA as a dimensionality reduction technique"
author: "Andrey Ziyatdinov"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples of using CPCA as a dimensionality reduction technique}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# About

## Include cpca package

```{r inc_cpca, cache = FALSE}
library(cpca)
```

## Include other packages

```{r inc, cache = FALSE}
library(plyr)
library(ggplot2)
library(gridExtra)

library(pander)

library(pls)
library(MASS)
```

Settings:

```{r settings, cache = FALSE, echo = T}
theme_set(theme_linedraw())
panderOptions('table.split.table', Inf)
panderOptions('knitr.auto.asis', FALSE)
```

# Examples

## Dataset iris

Loading the dataset:

```{r load_iris}
data(iris)
```

Preparing inputs to the CPCA model:

```{r prepare_iris}
X1 <- as.matrix(iris[, -5])
Y1 <- iris[, 5]
```

Running the CPCA model:

```{r iris_model}
m1 <- comprcomp(X1, Y1) 
```

Plotting scores:

```{r iris_varplot, fig.width = 9, fig.height = 3}
varplot(m1, X1, Y1)
```

```{r iris_scoreplot, fig.width = 6, fig.height = 4}
scoreplot(m1, X1, Y1)
```

## Bank notes dataset

Flury studied an interesting dataset known as the Swiss banknotes dataset,
which is available from an R package `mclust`.
The dataset consists of six measurements made on 200 banknotes,
where 100 are genuine and 100 are forged. 
The variables are: 

* width of the banknote; 
* height of the left-hand side of the banknote; 
* height of the right-hand side of the banknote; 
* distance between the top of the inner box to the upper border; 
* distance between the bottom of the inner box to the lower border; 
* diagonal of the inner box.


Loading the dataset:

```{r load_banknote}
data(banknote, package = "mclust")
```

```{r head_banknote, results = "asis"}
pander(head(banknote))
```

### Using all 6 variables

Preparing inputs to the CPCA model:

```{r prepare_banknote}
X2 <- as.matrix(banknote[, -1])
Y2 <- banknote[, 1]
```

Running the CPCA model:

```{r banknote_model}
m2 <- comprcomp(X2, Y2) 
```

Plotting scores:

```{r banknote_varplot, fig.width = 6, fig.height = 3}
varplot(m2, X2, Y2)
```

```{r banknote_scoreplot, fig.width = 6, fig.height = 4}
scoreplot(m2, X2, Y2)
```

The second component (CPC2, y-axis) on `scoreplot` captures little variance (5.7\%).
Let us see on the proportions of captured variance among CPCs (expressed in percentage).

```{r banknote_com_var, results = "asis"}
pander(varcomp(m2, X2, Y2, perc = TRUE))
```

#### Looking a higher-order components

```{r banknote_scoreplot_23, fig.width = 6, fig.height = 4}
scoreplot(m2, X2, Y2, comp = c(2, 3))
```

```{r banknote_scoreplot_34, fig.width = 6, fig.height = 4}
scoreplot(m2, X2, Y2, comp = c(3, 4))
```

#### PCA

```{r bankbote_pca, fig.width = 4, fig.height = 4}
m21 <- prcomp(X2)

scoreplot(m21, col = as.numeric(Y2) + 1)
```


