---
title: "Power algorithm for EVD"
author: "Andrey Ziyatdinov"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    toc: true
    keep_md: true
---

```{r options, echo = F}
opts_chunk$set(fig.path = "figures/", comment = NA, results = 'markup', tidy = F, message = F, warning = F, echo = T, cache = T)
```

```{r inc, cache = F, echo = F}
library(plyr)
library(reshape2)

library(ggplot2)
library(gridExtra)

library(pander)
```

```{r settings, cache = F, echo = F}
theme_set(theme_light())
panderOptions('table.split.table', Inf)
panderOptions('knitr.auto.asis', FALSE)
```

# About 

[wiki](https://en.wikipedia.org/wiki/Power_iteration): 

>In mathematics, the power iteration is an eigenvalue algorithm: given a matrix A, the algorithm will produce a number $\lambda$ (the eigenvalue) and a nonzero vector v (the eigenvector), such that $A v = \lambda v$. The algorithm is also known as the Von Mises iteration.

>The power iteration is a very simple algorithm. It does not compute a matrix decomposition, and hence it can be used when A is a very large sparse matrix. However, it will find only one eigenvalue (the one with the greatest absolute value) and it may converge only slowly.


## References

* [Blog post](http://blogs.sas.com/content/iml/2012/05/09/the-power-method.html) with SAS code (inspired this post)

# `eigenPower` function

```{r eigenPower}
eigenPower <- function(A, v0, tol = 1e-6, maxit = 1e3, 
  sparse = FALSE, sparseSymm = FALSE, 
  verbose = 0)
{
  timing <- list()
  timing$args <- proc.time()
  
  ### arguments
  stopifnot(!missing(A))
  
  if(missing(v0)) {
    v0 <- runif(ncol(A))
  }
  
  ### convert into sparse matrices
  if(any(sparse, sparseSymm)) {
    stopifnot(require(Matrix))
    
    if(sparse) {
      A <- Matrix(A, sparse = T)
    }
    
    if(sparseSymm) {
      A <- Matrix(A, sparse = T)
      A <- as(A, "symmetricMatrix")
    }
  }
  
  ### vars
  sparseMatrix <- FALSE
  cl <- attr(class(A), "package")
  if(!is.null(cl)) {
    if(cl == "Matrix") {
      sparseMatrix <- TRUE
    }
  }
  
  ### preparation before looping
  timing$algo <- proc.time()
  
  v0 <- as.numeric(v0)
  
  v <- v0
  v <- v / sqrt(v %*% v)

  ### loop
  lambda0 <- 0
  for(it in 1:maxit) {
    if(verbose > 1) {
      cat(" * it:", it, "/", maxit, "\n")
    }

    b <- tcrossprod(A, t(v)) # A %*% v
    v <- b / sqrt(as.numeric(crossprod(b)))
    lambda <- as.numeric(crossprod(v, b)) # t(v) %*% b
           
    delta <- abs((lambda - lambda0) / lambda)
    if(delta < tol) {
      break
    }
    
    lambda0 <- lambda
  }
  
  ### post-process
  converged <- (it < maxit)
  
  ### output
  timing$return <- proc.time()
  
  timing$targs <- timing$algo[["elapsed"]] - timing$args[["elapsed"]]
  timing$talgo <- timing$return[["elapsed"]] - timing$algo[["elapsed"]]
  
  out <- list(v0 = v0, tol = tol, maxit = maxit,
    it = it, delta = delta, converged = converged, 
    sparseMatrix = sparseMatrix,
    timing = timing,
    lambda = lambda, v = v)
  
  return(out)
}
```

## Test on a small example

The given example is borrowed from [here](http://blogs.sas.com/content/iml/2012/05/09/the-power-method.html).


```{r ex, dependson = -1} 
A <- matrix(c(-261, 209, -49, 
  -530, 422, -98,
  -800, 631, -144),
  ncol = 3, nrow = 3, byrow = TRUE)
  
v0 <- c(1, 2, 3)

out <- eigenPower(A, v0, maxit = 40, verbose = 2)
out[c("lambda", "v")]
```

```{r ex2, dependson = -2} 
out <- eigenPower(A, v0, maxit = 40, verbose = 2, sparse = TRUE)
out[c("lambda", "v")]
```

```{r eigen_A}
eigen(A)
```

## Test on CPU time


```{r input}
n <- seq(100, 500, by = 50)

out <- lapply(n, function(ni) {
  M <-  matrix(runif(ni * ni), ni, ni)

  lt <- lower.tri(M)
  ut <- upper.tri(M)

  M[lt] <- M[ut]
  
  list(n = ni,
    t.eigen = system.time(eigen(M))[["elapsed"]],
    t.eigenPower = system.time(eigenPower(M))[["elapsed"]])
})  
```

```{r plot_times, dependson = -1}
df <- ldply(out, function(x) data.frame(n = x$n, t_eigen = x$t.eigen, 
  t_eigenPower = x$t.eigenPower))

pf <- melt(df, id.vars = "n")

ggplot(pf, aes(n, value, color = variable)) + geom_point() + geom_line()

ggplot(pf, aes(n, value, color = variable)) + geom_point() + geom_line() + facet_wrap(~ variable, scales = "free_y")
```

## Test on CPU time (sparse matrices)

```{r test_sparse}
n <- seq(250, 1500, by = 250)

out <- lapply(n, function(ni) {
  M <-  matrix(runif(ni * ni), ni, ni)

  ind <- sample(seq(1, ni * ni), size = 0.85 * ni * ni, replace = F)
  M[ind] <- 0

  lt <- lower.tri(M)
  M[lower.tri(M)] = t(M)[lower.tri(M)]

  M1 <- Matrix(M, sparse = T)

  M2 <- Matrix(M, sparse = T)
  M2 <- as(M2, "symmetricMatrix")
    
  list(n = ni,
    t.eigenPower = system.time(eigenPower(M))[["elapsed"]],
    t.eigenPower.sparse = system.time(eigenPower(M, sparse = T))[["elapsed"]],
    t.eigenPower.sparseSymm = system.time(eigenPower(M, sparseSymm = T))[["elapsed"]])
})  
```


```{r plot_times2, dependson = -1}
df <- ldply(out, function(x) data.frame(n = x$n,
  t_eigenPower = x$t.eigenPower, 
  t_eigenPower_sparse = x$t.eigenPower.sparse,
  t_eigenPower_sparseSymm = x$t.eigenPower.sparseSymm))

pf <- melt(df, id.vars = "n")

ggplot(pf, aes(n, value, color = variable)) + geom_point() + geom_line()

ggplot(subset(pf, variable != "t_eigenPower"), aes(n, value, color = variable)) + geom_point() + geom_line()
```

## Test on CPU time (sparse matrices) #2

```{r test_sparse2}
n <- seq(250, 1500, by = 250)

out <- lapply(n, function(ni) {
  M <-  matrix(runif(ni * ni), ni, ni)

  ind <- sample(seq(1, ni * ni), size = 0.85 * ni * ni, replace = F)
  M[ind] <- 0

  lt <- lower.tri(M)
  M[lower.tri(M)] = t(M)[lower.tri(M)]

  
  out.eigenPower <- eigenPower(M)  
  out.eigenPower.sparse <- eigenPower(M, sparse = T)  
    
  list(n = ni,
    targs.eigenPower = out.eigenPower$timing$targs,
    talgo.eigenPower = out.eigenPower$timing$talgo,
    targs.eigenPower.sparse = out.eigenPower.sparse$timing$targs,
    talgo.eigenPower.sparse = out.eigenPower.sparse$timing$talgo    
  )
})  
```


```{r plot_times3, dependson = -1}
df <- ldply(out, function(x) data.frame(n = x$n,
  targs_eigenPower = x$targs.eigenPower, 
  talgo_eigenPower = x$talgo.eigenPower, 
  targs_eigenPower_sparse = x$targs.eigenPower.sparse, 
  talgo_eigenPower_sparse = x$talgo.eigenPower.sparse))  
  
pf <- melt(df, id.vars = "n")
pf <- mutate(pf, t = substr(variable, 1, 5))

ggplot(pf, aes(n, value, color = variable)) + geom_point() + geom_line()

ggplot(subset(pf, t == "targs"), aes(n, value, color = variable)) + geom_point() + geom_line()

ggplot(subset(pf, t == "talgo"), aes(n, value, color = variable)) + geom_point() + geom_line()
```
